---
title: Evolution of R Charting of Time Series
subtitle: from plot.default to Interactive rCharts
author: Timely Portfolio
github: {user: ramnathv, repo: rCharts, branch: "gh-pages"}
framework: minimal
mode: selfcontained
ext_widgets: {rCharts: ["libraries/morris","libraries/nvd3", "libraries/polycharts", "libraries/highcharts","libraries/xcharts", "libraries/leaflet", "libraries/rickshaw"]}
hitheme: solarized_light
logo: libraries/frameworks/minimal/images/rCharts.png
---

<style>
.rChart {
  height: 400px
}
</style>

# Evolution of R Charting of Time Series

As with all of R, the ability to easily chart financial time series is a result of an iterative evolution driven by the collaboration of an extremely dedicated group of open source volunteers.  With the release of [rCharts](http://ramnathv.github.io/rCharts), I thought it would be interesting to document the steps in the evolution.  For each step, we will include a link to the source (svn or github) and a minimal example to demo the "out-of-the-box" capability.  Just in case you don't make it to the end,   
> Thanks to the contributors!  I wouldn't be using R if it weren't for you.

```{r echo = F, message = F, cache = F}
require(rCharts)
options(RCHART_WIDTH = 600, RCHART_HEIGHT = 400)
knitr::opts_chunk$set(
  comment = NA,
  results = 'asis',
  tidy = F,
  error = F,
  message = F,
  warning = F,
  fig.width = 8,
  fig.height = 5)
```

---
## Plots Need Data
First, to build a plot, we need data. Let's show how easy it is to get a time series of financial data in R through `quantmod` [`getSymbols()`](https://r-forge.r-project.org/scm/viewvc.php/pkg/R/getSymbols.R?root=quantmod&view=log).  The `getSymbols()` function has been a work in progress since December 20, 2006.

```{r}
require(latticeExtra)
require(quantmod)
require(reshape2)
require(rCharts)
require(PerformanceAnalytics)
require(xtsExtra)


# get S&P 500 data from FRED (St. Louis Fed)
sp500 <- na.omit( 
  getSymbols(
    "SP500",
    src = "FRED",
    from = "1949-12-31",
    auto.assign = FALSE
  )
)

# use monthly data
sp500.monthly <- sp500[endpoints(sp500, on ="months")]
```

---
## Steps in the Evolution

### plot.default (As Old as Time Itself)

```{r}
# base plot of time series prior to xts
# get the data in data.frame format rather than xts
sp500.df <- data.frame(
  index(sp500.monthly),
  coredata(sp500.monthly),
  stringsAsFactors=FALSE
)
# name columns
colnames( sp500.df ) <- c( "date", "sp500" )
# go back in time to plot.default from the graphics library
graphics::plot.default(
  x = sp500.df$date,
  y = sp500.df$sp500,
  type = "l",
  xlab = "Date",
  ylab = "Closing Value",
  main = "S&P 500 (graphics::plot.default)"
)
```

- - -

---
### [zoo](https://r-forge.r-project.org/scm/viewvc.php/pkg/zoo/R/plot.zoo.R?root=zoo&view=log) 2004-10-08


```{r}
# 2004-10-08 plot.zoo comes to the rescue with the zoo package
zoo::plot.zoo(sp500.monthly)
```

- - -

---
### [zoo](https://r-forge.r-project.org/scm/viewvc.php/pkg/zoo/R/xyplot.zoo.R?root=zoo&view=log) Meets [lattice](http://r-forge.r-project.org/scm/?group_id=638) 2006-07-06

```{r}
# 2006-07-06 zoo also made a handy lattice xyplot.zoo function
# so no more need to melt wide data to long data prior to plotting
# although slightly out of chronology
# I'll also use theEconomist from latticeExtra
asTheEconomist(
  xyplot(
    sp500.monthly,
    scales = list( y = list( rot = 0 ) ),
    main = "S&P 500 (lattice::xyplot.xts)"  
  )
)
```

- - -

---
### [PerformanceAnalytics chart.TimeSeries](https://r-forge.r-project.org/scm/viewvc.php/pkg/PerformanceAnalytics/R/chart.TimeSeries.R?root=returnanalytics&view=log) 2007-02-02

```{r}
# 2007-02-02 chart.TimeSeries in PerformanceAnalytics
charts.PerformanceSummary(
  ROC(sp500.monthly, n = 1, type = "discrete"),
  main = "S&P 500 (PerformanceAnalytice::charts.PerformanceSummary)"
)
```

- - -

---
### [quantmod/ttr chartSeries](https://r-forge.r-project.org/scm/viewvc.php/pkg/R/chartSeries.R?root=quantmod&view=log) 2007-10-07
```{r}
# 2007-10-17 then quantmod/ttr built on zoo
# to offer much better handling of financial time series
# notice the ease of adding pertinent financial information
chartSeries(
  sp500.monthly,
#  log = TRUE,
  theme = chartTheme("white"),
  TA = c(addBBands(),addTA(RSI(sp500.monthly)))
)
```

Just look how easy it is to zoom.

```{r}
# also easy zooming
zoomChart("1990::")
```

- - -

---
### [xts plot.xts](https://r-forge.r-project.org/scm/viewvc.php/pkg/xts/R/plot.R?root=xts&view=log) 2008-02-17

```{r}
# 2008-02-17 xts improved zoo and other time series libraries
# http://cran.r-project.org/web/packages/xts/vignettes/xts.pdf
# plot.zoo got ported to plot.xts and little graphing improvement
xts::plot.xts(
  sp500.monthly,
  ylab = "Closing Value",
  main = "S&P 500 (xts::plot.xts)"  
)
```

- - -

---
### [xtsExtra plot.xts and barplot.xts](https://r-forge.r-project.org/scm/viewvc.php/pkg/xtsExtra/R/plot.R?root=xts&view=log) 2012-05-30

```{r}
# Summer 2012 Google Summer of Code
# xtsExtra significantly improves xts
# lots of examples in this post
# http://timelyportfolio.blogspot.com/search/label/plot.xts

#require(devtools)
#setwd("C:/Program Files/R/R-2.15.1/sandbox/svnsource/xts/pkg/xtsExtra")
#build()
#load_all()


#explore barplot.xts to do a chart of annual returns for both indexes
#merge prices
getSymbols("^RUT", from = "1900-01-01")  
prices <- merge(sp500,RUT[,4])
#use endpoints to get annual returns
returns.annual <- as.xts(
  apply(
    ROC(prices[endpoints(prices,"years")],type="discrete",n=1),
    MARGIN = 2,
    FUN = na.fill, fill = 0
  ),
  order.by = index(prices[endpoints(prices,"years")])
)
#name columns something a little more clear
colnames(returns.annual) <- c("S&P 500","Russell 2000")
#using barplot.xts create the plot
#I made some subtle changes to barplot.xts to experiment so plot will be cosmetically different
barplot.xts(returns.annual,
            stacked=FALSE,
            box="transparent",  #get rid of box surrounding the plot
            ylim=c(-0.5,0.5),
            ylab=NA,
            border=c(brewer.pal(n=11,"BrBG")[c(4,9)]),
            col=c(brewer.pal(n=11,"BrBG")[c(4,9)])) #deliberately trying some new colors                         
title(main="Annual Returns of S&P 500 and Russell 2000", 
      outer = TRUE,
      adj=0.05, font.main = 1, cex.main = 1.25, line = -2)
```

- - -

---
### rCharts

```{r}
# 2013 the world changes with rCharts

# define a function to convert wide to long
xtsMelt <- function(xtsData,metric){
  df <- data.frame(index(xtsData),coredata(xtsData),stringsAsFactors=FALSE)
  df.melt <- melt(df,id.vars=1)
  df.melt <- data.frame(df.melt,rep(metric,NROW(df.melt)))
  #little unnecessary housekeeping
  df.melt <- df.melt[,c(1,2,4,3)]
  colnames(df.melt) <- c("date","indexname","metric","value")
  df.melt$date <- as.Date(df.melt$date)
  #javascript works better when there are no .
  #remove troublesome . using modified method from this Stack Overflow
  #http://stackoverflow.com/questions/2851015/convert-data-frame-columns-from-factors-to-characters
  i <- sapply(df.melt, is.factor)
  df.melt[i] <- lapply(df.melt[i], gsub, pattern="\\.", replacement="")
  
  return(df.melt)
}

sp500.melt <- xtsMelt(
  sp500.monthly,
  metric = "price"
)

n1 <- nPlot(
  value~date,
  data = sp500.melt,
  group = "indexname",  # even though only one series need to specify group
  type = "lineWithFocusChart"
)

n1$xAxis(
  tickFormat="#!function(d) {return d3.time.format('%b %Y')(new Date( d * 86400000 ));}!#"
)

n1$x2Axis(
  tickFormat="#!function(d) {return d3.time.format('%Y')(new Date( d * 86400000 ));}!#"
)
n1$print("chart1")
```



<h4>morris.js example</h4>

<h4>rickshaw example</h4>


- - -

---
## Thanks
Thanks to all the wonderful and diligent contributors who have made R great.
