---
title: R Time Series Plotting
subtitle: from plot.default to rCharts
author: Timely Portfolio
github: {user: timelyportfolio, repo: rCharts_time_series, branch: "gh-pages"}
framework: minimal
mode: selfcontained
ext_widgets: {rCharts: ["libraries/morris","libraries/nvd3", "libraries/polycharts", "libraries/highcharts","libraries/xcharts", "libraries/rickshaw"]}
hitheme: solarized_light
---

<style>
.rChart {
  height: 400px
}
</style>

# History of R Charting of Time Series

As with all of R, the ability to easily chart financial time series is the result of an iterative progression driven by the collaboration of an extremely dedicated group of open source volunteers.  With the release of [rCharts](http://ramnathv.github.io/rCharts), I thought it would be interesting to document the timeline of this progression.  For each step in the timeline, I will include a link to the source code (svn or github) of the package and a minimal example to demo the "out-of-the-box" capability.  Separating the financial time series piece from graphing in general can get murky, and some of the timeline will differ from the timeline of R graphics and the timeline of R time series analysis.

For a much more extensive discussion of time series analysis with R, please see:

- [Time Series Analysis with R](http://www.stats.uwo.ca/faculty/aim/tsar/tsar.pdf) by A. Ian McLeod, Hao Yu, and Esam Mahdi
- [CRAN Task View: Time Series Analysis](http://cran.r-project.org/web/views/TimeSeries.html) by Rob Hyndman
- [A Little Book of R for Time Series](http://a-little-book-of-r-for-time-series.readthedocs.org/en/latest/src/timeseries.html) by Avril Chohlan

Just in case you don't make it to the end,   
> Thanks to the contributors!  I wouldn't be using R if it weren't for you.

```{r echo = F, message = F, cache = F}
options(RCHART_WIDTH = 600, RCHART_HEIGHT = 400)
knitr::opts_chunk$set(
  comment = NA,
  results = 'asis',
  tidy = F,
  error = T,
  message = F,
  warning = F,
  fig.width = 8,
  fig.height = 5,
  fig.keep = 'last')
```
First, to build a plot, we need data. Let's see how easy it is to get a time series of financial data in R through `quantmod` [`getSymbols()`](https://r-forge.r-project.org/scm/viewvc.php/pkg/R/getSymbols.R?root=quantmod&view=log).  The [`getSymbols()`]((https://r-forge.r-project.org/scm/viewvc.php/pkg/R/getSymbols.R?root=quantmod&view=log)) function has been a work in progress since December 20, 2006.

```{r}
require(latticeExtra)
require(ggplot2)
require(reshape2)
require(quantmod)
require(PerformanceAnalytics)
require(xtsExtra)
require(rCharts)


# get S&P 500 data from FRED (St. Louis Fed)
sp500 <- na.omit( 
  getSymbols(
    "SP500",
    src = "FRED",
    from = "1949-12-31",
    auto.assign = FALSE
  )
)

# use monthly data
sp500.monthly <- sp500[endpoints(sp500, on ="months")]
```

---
## Timeline

### plot.default (As Old as Time Itself)

```{r}
# base plot of time series prior to xts
# get the data in data.frame format rather than xts
sp500.df <- data.frame(
  index(sp500.monthly),
  coredata(sp500.monthly),
  stringsAsFactors=FALSE
)
# name columns
colnames( sp500.df ) <- c( "date", "sp500" )
# go back in time to plot.default from the graphics library
graphics::plot.default(
  x = sp500.df$date,
  y = sp500.df$sp500,
  type = "l",
  xlab = "Date",
  ylab = "Closing Value",
  main = "S&P 500 (graphics::plot.default)"
)
```

- - -

---
### [ts](https://stat.ethz.ch/pipermail/r-announce/1999/000097.html) 1999-08-27
The ts package was added in R version 0.65.0 and significantly improved with release 1.5.0 in April 2002.  There is a very good discussion of the improvements in Brian Ripley's ["Time Series in R 1.5.0" from Volume 2 of R News, June 2002](http://cran.r-project.org/doc/Rnews/Rnews_2002-2.pdf).  plot.ts added some nice features, such as the ability to handle multiple series, specify panels per series, and automatically handle acf, ARIMA, and HoltWinters.

```{r}
stats::plot.ts(
  ts(sp500.monthly,
     start = c(
       as.numeric(format(index(sp500.monthly)[1],"%Y")),
       as.numeric(format(index(sp500.monthly)[1],"%m"))
     ),
     frequency = 12
  ), # some backwards conversion to ts from xts
  xlab = "Date",
  ylab = "Closing Value",
  main = "S&P 500 (stats::plot.ts)"
)
```

- - -

---
### [lattice](http://r-forge.r-project.org/scm/?group_id=638) and [grid](http://www.stat.auckland.ac.nz/~paul/grid/grid.html) released with R 1.5.0 2002-04-29
With the release of lattice and grid and also the improvements in ts mentioned above, R 1.5.0 was a very important milestone for both graphing and time series analysis.  All of these are covered in [Volume 2 of R News, June 2002](http://cran.r-project.org/doc/Rnews/Rnews_2002-2.pdf).  lattice began the era of aesthetically pleasing and production-quality graphics straight from R.

- - -

---
### [zoo](https://r-forge.r-project.org/scm/viewvc.php/pkg/zoo/R/plot.zoo.R?root=zoo&view=log) 2004-10-08


```{r}
# 2004-10-08 plot.zoo comes to the rescue with the zoo package
zoo::plot.zoo(
  sp500.monthly,
  main = "S&P 500 (zoo::plot.zoo)"
)
```

- - -

---
### [zoo](https://r-forge.r-project.org/scm/viewvc.php/pkg/zoo/R/xyplot.zoo.R?root=zoo&view=log) Meets [lattice](http://r-forge.r-project.org/scm/?group_id=638) 2006-07-06

```{r}
# 2006-07-06 zoo also made a handy lattice xyplot.zoo function
# so no more need to melt wide data to long data
# or convert to data.frame prior to plotting
# although slightly out of chronology
# I'll also use theEconomist from latticeExtra
asTheEconomist(
  xyplot(
    sp500.monthly,
    scales = list( y = list( rot = 0 ) ),
    main = "S&P 500 (lattice::xyplot.xts)"  
  )
)
```

- - -

---
### [PerformanceAnalytics chart.TimeSeries](https://r-forge.r-project.org/scm/viewvc.php/pkg/PerformanceAnalytics/R/chart.TimeSeries.R?root=returnanalytics&view=log) 2007-02-02

```{r}
# 2007-02-02 chart.TimeSeries in PerformanceAnalytics
charts.PerformanceSummary(
  ROC(sp500.monthly, n = 1, type = "discrete"),
  main = "S&P 500 (PerformanceAnalytice::charts.PerformanceSummary)"
)
```

- - -

---
### [ggplot2](http://cran.r-project.org/src/contrib/Archive/ggplot2/) 2007-06-10
Although ggplot2 is not designed specifically for time series plotting, I include it in the timeline for both its significant impact on R graphics and its ability to handle dates/times on the x-axis.  To use xts with ggplot2, a simple conversion to a wide or long format data.frame is necessary.

```{r}
#ggplot2 requires conversion of xts to data.frame
#we will use the data.frame from the plot.default example
ggplot( sp500.df, aes(date) ) + 
  geom_line( aes( y = sp500 ) ) +
  labs( title = "S&P 500 (ggplot2::ggplot)")
```

- - -

---
### [quantmod/TTR chartSeries](https://r-forge.r-project.org/scm/viewvc.php/pkg/R/chartSeries.R?root=quantmod&view=log) 2007-10-07

```{r}
# 2007-10-17 then quantmod/TTR built on zoo
# to offer much better handling of financial time series
# notice the ease of adding pertinent financial information
chartSeries(
  sp500.monthly,
#  log = TRUE,
  theme = chartTheme("white"),
  TA = c(addBBands(),addTA(RSI(sp500.monthly)))
)
```

Just look how easy it is to zoom.

```{r}
# also easy zooming
zoomChart("1990::")
```

- - -

---
### [xts plot.xts](https://r-forge.r-project.org/scm/viewvc.php/pkg/xts/R/plot.R?root=xts&view=log) 2008-02-17

```{r}
# 2008-02-17 xts improved zoo and other time series libraries
# http://cran.r-project.org/web/packages/xts/vignettes/xts.pdf
# plot.zoo got ported to plot.xts and little graphing improvement
xts::plot.xts(
  sp500.monthly,
  ylab = "Closing Value",
  main = "S&P 500 (xts::plot.xts)"  
)
```

- - -

---
### [timeSeries plot](https://r-forge.r-project.org/scm/viewvc.php/pkg/timeSeries/R/methods-plot.R?root=rmetrics&view=log) 2009-05-17
The timeSeries plot method is basically a port of R's plot.ts().  It does not significantly add any plotting functionality, but I include it for completeness and since the [Rmetrics team](https://www.rmetrics.org) offers robust financial analysis through its many R packages that depend on the timeSeries object.

```{r}
require(timeSeries)
timeSeries::plot(
  timeSeries(sp500.monthly),
  main = "S&P 500 (timeseries::plot)"
)
```

- - -

---
### [xtsExtra plot.xts and barplot.xts](https://r-forge.r-project.org/scm/viewvc.php/pkg/xtsExtra/R/plot.R?root=xts&view=log) 2012-05-30

```{r}
# Summer 2012 Google Summer of Code
# xtsExtra significantly improves xts
# lots of examples in this post
# http://timelyportfolio.blogspot.com/search/label/plot.xts

#require(devtools)
#setwd("C:/Program Files/R/R-2.15.1/sandbox/svnsource/xts/pkg/xtsExtra")
#build()
#load_all()


#explore barplot.xts to do a chart of annual returns for both indexes
#merge prices
russell2000 <- getSymbols("^RUT", from = "1900-01-01", auto.assign = F)  
prices <- merge(sp500,russell2000[,4])
#use endpoints to get annual returns
returns.annual <- as.xts(
  apply(
    ROC(prices[endpoints(prices,"years")],type="discrete",n=1),
    MARGIN = 2,
    FUN = na.fill, fill = 0
  ),
  order.by = index(prices[endpoints(prices,"years")])
)
#name columns something a little more clear
colnames(returns.annual) <- c("S&P 500","Russell 2000")

barplot.xts(
  returns.annual,
  stacked=FALSE,
  box="transparent",  #get rid of box surrounding the plot
  ylim=c(-0.5,0.5),
  ylab=NA,
  border=c(brewer.pal(n=11,"BrBG")[c(4,9)]),
  col=c(brewer.pal(n=11,"BrBG")[c(4,9)])
)

title(
  main="Annual Returns of S&P 500 and Russell 2000 (xtsExtra::barplot.xts)", 
  outer = TRUE,
  adj=0.05,
  font.main = 1,
  cex.main = 1.25,
  line = -2
)
```

- - -

---
### [rCharts](http://rcharts.github.io/site) 2013

```{r}
# 2013 the world changes with rCharts

# define a function to convert xts wide to long data.frame
xtsMelt <- function(xtsData,metric){
  df <- data.frame(index(xtsData),coredata(xtsData),stringsAsFactors=FALSE)
  df.melt <- melt(df,id.vars=1)
  df.melt <- data.frame(df.melt,rep(metric,NROW(df.melt)))
  #little unnecessary housekeeping
  df.melt <- df.melt[,c(1,2,4,3)]
  colnames(df.melt) <- c("date","indexname","metric","value")
  df.melt$date <- as.Date(df.melt$date)
  #javascript works better when there are no .
  #remove troublesome . using modified method from Stack Overflow
  i <- sapply(df.melt, is.factor)
  df.melt[i] <- lapply(df.melt[i], gsub, pattern="\\.", replacement="")
  
  return(df.melt)
}

sp500.melt <- xtsMelt(
  sp500.monthly,
  metric = "price"
)

n1 <- nPlot(
  value~date,
  data = sp500.melt,
  group = "indexname",  # even though only one series need to specify group
  type = "lineWithFocusChart"
)

n1$xAxis(
  tickFormat=
    "#!function(d) {return d3.time.format('%b %Y')(new Date( d * 86400000 ));}!#"
)

n1$x2Axis(
  tickFormat=
    "#!function(d) {return d3.time.format('%Y')(new Date( d * 86400000 ));}!#"
)
n1$print("chart1")
```



<h4>morris.js example</h4>

<h4>rickshaw example</h4>

<h4>highcharts example</h4>


- - -

---
## Thanks
Thanks to all the wonderful and diligent contributors who have made R great.


## References
http://www.stats.uwo.ca/faculty/aim/tsar/tsar.pdf

http://cran.r-project.org/web/views/TimeSeries.html

http://cran.r-project.org/web/packages/zoo/vignettes/zoo.pdf

http://cran.r-project.org/web/packages/timeSeries/timeSeries.pdf

http://cran.r-project.org/web/packages/xts/vignettes/xts.pdf




http://fishyoperations.com/r/timeline-graph-with-ggplot2/
